Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2021-06-23 paper/tidy-missing-data-paper.Rmd
misc updates to paper to prepare for new merge

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is 'tidy data,' which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline [@Wickham2014]. Another idea is the grammar of graphics, describing how to map data values into a visualization [@Wilkinson2012]. These principles have been widely adopted, but do not address the problem of missing data. In particular, there is little guidance on how to handle missing values in a data analysis workflow. Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).

@@

p1 + grid::textGrob(
  label = "Warning message:
Removed 171 rows containing 
missing values (geom_point).", 
gp = grid::gpar(fontsize = 14,
                fontfamily = "Courier",
                col = "salmon", 
                fontface = "bold"))

@@

> 1.  Each variable must have its own column.
> 2.  Each observation must have its own row.
> 3.  Each value must have its own cell.

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

There are many ways each value can be missing, we adopt the notation used in [@vanBuuren2018]. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the *missing data model* $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "\\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

@@

2.  **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>". For example: "NA_instrument" uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).

3.  **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var". This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state. Although this might break naming conventions for lower case "snake_case" variables (e.g., @Wickham2014), this distinguishes these columns from other columns.

4.  **Connectedness**: Binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).

@@

```{r nabularfig, echo = FALSE, fig.cap = "The process of creating nabular data. Data transformed to shadow matrix, where values are either not missing or missing: '!NA' or 'NA'. The shadow matrix can be converted to long form to create missingness summary plots. Nabular data is created by binding the columns of the data and shadow matrix. Special missing values (such as -99) are identified as special missings, and values imputed and tracked. Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.", out.width = "100%", fig.align = "center"}

@@

*Nabular* data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn]. Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \@ref(verbs)).

@@

### scan: Searching for common missing value labels {#verbs-search}

@@

```{r miss-scan-count, eval = FALSE}
miss_scan_count(data = dat_ms, 
                search = -99) %>% 
  knitr::kable(
    caption = "Table of the occurrences of the search '-99' in the data, 'dat\\_ms'. There is one occurrence if -99 in variables x and z.")
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

### replace: Replacing values with missing values {#verbs-replace-with}

@@

### add: Adding missingness summary variables {#verbs-add-cols}

@@

tibble::tibble(
  `Function` = c(
    "add_n_miss(data)",
    "add_prop_miss(data)",
    "add_miss_cluster(data)"
    ),
  `Adds column which:` = c(
    "Contains the number missing values in a row",
    "Contains the proportion of missing values in a row",
    "Contains the missing value cluster"
  )) %>% 
    knitr::kable(caption = "Overview of the 'add' functions in naniar")

@@

### shadow: Creating nabular data {#verbs-nabular}

@@

### flag: Describing different types of missing values {#verbs-recode}

@@

### impute: Imputing values {#verbs-impute}

@@

### track: Shadow and impute missing values {#verbs-track}

@@

```{r impute-summary-out, echo = FALSE}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max))  %>% 
  kable(caption = "Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable 'Ozone'. The 'any\\_missing' column denotes imputed values ('Missing', since they were previously missing), and non-imputed values ('Not Missing'). The mean and median values are similar, but the minimum and maximum values are very different.")
```

@@

Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: "Land size", "latitude", "longitude", "bedroom2", "bathroom", "car", and "land size".

@@

<!-- Note: I think I'm basically there, but I'd like to revisit the first paragraph in this section and the last sentence -->

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
    includes:
      in_header: "preamble.tex"
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is "tidy data", which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline [@Wickham2014]. Another idea is the grammar of graphics, describing how to map data values into a visualization [@Wilkinson2012]. These principles have been widely adopted, but do not address the problem of missing data. In particular, there is little guidance on how to handle missing values in a data analysis workflow. Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).

@@

p1 + grid::textGrob(
  label = "Warning message:
removed 171 rows containing 
missing values (geom_point).", 
gp = grid::gpar(fontsize = 14,
                fontfamily = "Courier",
                col = "salmon", 
                fontface = "bold"))

@@

> 1.  each variable must have its own column.
> 2.  each observation must have its own row.
> 3.  each value must have its own cell.

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

There are many ways each value can be missing, we adopt the notation used in [@vanBuuren2018]. The information in $B$ can be used to arrive at three categories of missing values: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the *missing data model* $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = '\\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional "NA" boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.', fig.show = "asis"}

@@

2.  **Special missing values**: building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>". For example: "NA_instrument" uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).

3.  **Coordinated names**: variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var". This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state. Although this might break naming conventions for lower case "snake_case" variables (e.g., @Wickham2014), this distinguishes these columns from other columns.

4.  **Connectedness**: binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).

@@

```{r nabularfig, echo = FALSE, fig.cap = 'The process of creating nabular data. Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA". The shadow matrix can be converted to long form to create missingness summary plots. Nabular data is created by binding the columns of the data and shadow matrix. Special missing values (such as -99) are identified as special missings, and values imputed and tracked. Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.', out.width = "100%", fig.align = "center"}

@@

*Nabular* data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn]. Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) keep one thing in a cell and (2) describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \@ref(verbs)).

@@

### scan: searching for common missing value labels {#verbs-search}

@@

```{r miss-scan-count, eval = FALSE}
miss_scan_count(data = dat_ms, 
                search = -99) %>% 
  knitr::kable(
    caption = 'Table of the occurrences of the search "-99" in the data, "dat\\_ms". There is one occurrence if -99 in variables x and z.')
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))
```

### replace: replacing values with missing values {#verbs-replace-with}

@@

### add: adding missingness summary variables {#verbs-add-cols}

@@

tibble::tibble(
  `Function` = c(
    "add_n_miss(data)",
    "add_prop_miss(data)",
    "add_miss_cluster(data)"
    ),
  `Adds column which:` = c(
    "contains the number missing values in a row",
    "contains the proportion of missing values in a row",
    "contains the missing value cluster"
  )) %>% 
    knitr::kable(caption = 'Overview of the "add" functions in naniar')

@@

### shadow: creating nabular data {#verbs-nabular}

@@

### flag: describing different types of missing values {#verbs-recode}

@@

### impute: imputing values {#verbs-impute}

@@

### track: shadow and impute missing values {#verbs-track}

@@

```{r impute-summary-out, echo = FALSE}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max))  %>% 
  kable(caption = 'Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone". The "any\\_missing" column denotes imputed values ("Missing", since they were previously missing), and non-imputed values ("Not Missing"). The mean and median values are similar, but the minimum and maximum values are very different.')
```

@@

Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: "latitude", "longitude", "bedroom2", "bathroom", "car", and "land size".

@@
